The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Natural or tidal?

### 2025-09-27 thesis/results.tex
a compltet thesis draft

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
\caption[Sculptor simulation snapshots]{Images of the dark matter
evolution over a selection of past apocentres and the present day.
Limits range from -150 to 150 kpc in the \(y\)--\(z\) (approximately
orbital) plane and the colourscale is logarithmic spanning 5 orders of
magnitude between the maximum and minimum values. The green dot
represents the final expected position of the galaxy and the solid and
dotted grey curves represents the orbit over one previous or future
radial oscillation respectively.}\label{fig:scl_sim_images}
\end{figure}

@@

\subsection{Robustness to initial
conditions}\label{robustness-to-initial-conditions}

Some initial conditions may be slightly more susceptible to tides. Dark
matter halos which are cored or less concentrated disrupt or lose mass
faster \citep[e.g.,][]{stucker+2023}. We consider alternative dark
matter halo initial conditions in the Appendix. We find that while the
details of the strength of tidal impacts depend on the halo, our models
here remain representative of the tidal effects.

@@

The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
case. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM evolution.
UMi loses more DM mass, with only about 3\% of the total mass remaining
({[}Table~\ref{tbl:umi_sim_results}; {]}). However, tidal features in
the observed stellar components are still extremely faint, appearing
outside 100 arcminutes in Fig.~\ref{fig:umi_smallperi_i_f}. Tides only
moderately affect the observed size and velocity dispersion of Ursa
Minor (Table~\ref{tbl:umi_sim_stars_results}).

@@

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_plummer_i_f.pdf}}
\caption[Ursa Minor simulated density profiles]{The tidal effects on the
stellar surface density of Ursa Minor for Plummer stars on the
\smallperi{} orbit.}\label{fig:umi_plummer_i_f}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_stream.pdf}}
\caption[Ursa Minor Predicted stream]{The properties of the stream
around the UMi \smallperi{} orbit with Plummer stars. The panels are all
as a function of \(\xi'\), the distance along the stream as defined by
the current GSR proper motion vector. The top panels show the GSR proper
motions in RA and Dec, and the bottom two show the distance and GSR
radial velocities. To sample the stream, we randomly draw 100,000
samples from the snapshot based on the stellar weights. A detectible
gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
the stream is tracked across several
degrees.}\label{fig:umi_tidal_stream}
\end{figure}

@@

>>>>

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_sim_images.png}}
\caption[Sculptor simulation snapshots]{Images of the dark matter
evolution over a selection of past apocentres and the present day.
Limits range from -150 to 150 kpc in the \(y\)--\(z\) (\(\sim\)orbital)
plane, and the colourscale is logarithmic spanning 5 orders of magnitude
between the maximum and minimum values. The green dot represents the
final expected position of the galaxy and the solid and dotted grey
curves represents the orbit over one previous or future radial
oscillation respectively.}\label{fig:scl_sim_images}
\end{figure}

@@

@@

The tidal evolution of Ursa Minor is similar to Sculptor in the MW-only
case. Fig.~\ref{fig:umi_sim_images} shows snapshots of the DM evolution.
UMi loses more DM mass, with only about 3\% of the total mass remaining
(see Fig.~\ref{fig:umi_tidal_track} and
Table~\ref{tbl:umi_sim_results}). However, tidal features in the
observed stellar components are still extremely faint, appearing outside
100 arcminutes in Fig.~\ref{fig:umi_smallperi_i_f}. Tides only
moderately affect the observed size and velocity dispersion of Ursa
Minor (Table~\ref{tbl:umi_sim_stars_results}).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_tidal_track.png}}
\caption[Ursa Minor tidal tracks]{Similar to
Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
substantially more mass than Sculptor.}\label{fig:umi_tidal_track}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_plummer_i_f.pdf}}
\caption[Ursa Minor Plummer model density]{The tidal effects on the
stellar surface density of Ursa Minor for Plummer stars on the
\smallperi{} orbit.}\label{fig:umi_plummer_i_f}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_sim_stream.pdf}}
\caption[Ursa Minor predicted stream]{The properties of the stream
around the UMi \smallperi{} orbit with Plummer stars. The panels are all
as a function of \(\xi'\), the distance along the stream as defined by
the current GSR proper motion vector. The top panels show the GSR proper
motions in RA and Dec, and the bottom two show the distance and GSR
radial velocities. To sample the stream, we randomly draw 100,000
samples from the snapshot based on the stellar weights. A detectible
gradient in \(\mu_{\alpha*}\) and LOS velocity should be detectible if
the stream is tracked across several
degrees.}\label{fig:umi_tidal_stream}
\end{figure}

@@

\section{Modeling uncertainties}\label{modeling-uncertainties}

\subsection{Modelling assumptions}\label{modelling-assumptions}

As Section~\ref{sec:results} shows, tides marginally affected the
stellar components of Scl and UMi, even under extreme orbits. While
Section~\ref{sec:results} only presents select models, alternative
initial conditions do not affect our qualitative conclusions on tidal
effects.

We neglect baryonic physics. Since Scl and UMi have predominantly stars
older than \(\sim 9\) Gyr
\citep{carrera+2002, deboer+2011, weisz+2014, delosreyes+2022, sato+2025},
gas dynamics are unlikely to affect recent evolution. A collisionless
dark-matter only simulation should be an excellent approximation.

Cored or less concentrated dark matter halos disrupt quicker
\citep[e.g.,][]{stucker+2023}. Our fiducial UMi halo in particular is
among the least concentrated halos consistent with UMi's velocity
dispersion. Although Scl's fiducial halo is more concentrated, less
concentrated and cored halos evolve similarly (see
Section~\ref{sec:extra_results}).

Galaxies are rarely perfect isotropic spheres. Sculptor and Ursa Minor
are elliptical, and halos are expected to be radially anisotropic
\citep[e.g.,][]{navarro+2010}. We test example non-spherical and
anisotropic models in Section~\ref{sec:extra_results}, finding that
these assumptions likely do not alter our conclusions.

A system's observed velocity dispersion directly constrains the mean
density within \(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force
required to disrupt the stellar component does not strongly depend on
the inner halo structure (via the Jacobi radius). Alternative initial
conditions may influence the total mass evolution but would produce a
similar final stellar structure.

\subsection{Orbital uncertainties}\label{orbital-uncertainties}

The long-term orbits of satellites are uncertain. Analytic Milky Way
potentials neglect many unknown details, including triaxiality, halo
twisting, mass evolution, and substructure. As a result, calculated
orbits may diverge significantly from the true orbits of satellites
\citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the Milky Way and
dynamical friction implies that orbits are typically less bound in the
past (less affected by tides). However, orbital energy and angular
momentum are not conserved in cosmological simulations. Orbits in
analytic potentials may overestimate the pericentre and underestimate
the maximum tidal stress \citep[although typically not by enough to
change our conclusions,][]{santistevan+2023, santistevan+2024}.

Fig.~\ref{fig:scl_orbit_lmc_uncert} illustrates the uncertainties of the
long-term orbits of Scl and UMi. Scl distance could range from
4--\(300\,\kpc\) any time more than 4Gyr ago. And, if the LMC had a
previous pericentre, UMi may have been an LMC satellite
\citep[see][]{vasiliev2024}. Notably, these potentials are still
idealized in many ways.

Motivated by Fig.~\ref{fig:scl_orbit_lmc_uncert}, we examine an extreme
pericentre of \(4\,\kpc\) of Scl with the MW in Appendix
\ref{sec:extra_results}, finding it still insufficient to produce the
observed density profile. We conclude our simulated orbits represent
reasonable extremes for \emph{recent} tidal effects. Past encounters
with the LMC are revisited below as a form of ``pre-processing'' in
Section~\ref{sec:stellar_halos}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_lmc_orbits_mass_effect.png}}
\caption[Sculptor Orbits with LMC]{The long-term orbital history of
Sculptor (\textbf{top}) and UMi (\textbf{bottom}) are uncertain. In both
panels, light, transparent lines represent randomly-sampled orbit of the
satellites (alla ref) in three different LMC/MW mass models from
\citet{vasiliev2024}. The LMC orbits are in solid, thick lines of the
corresponding colour. The L2M11 has a lighter LMC mass, and the L3M10
model has a lighter MW mass than our fiducial L3M11 LMC
model.}\label{fig:scl_orbit_lmc_uncert}
\end{figure}

\subsection{Summary}\label{summary-2}

While the long-term tidal evolution is unconstrained, we conclude that
our models are reasonable representations of recent tidal effects. As a
result, recent tides are unlikely to affect the stellar distributions of
Sculptor or Ursa Minor.
